\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2014-09-05 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
P3 description

One area that has been very effective at reducing the number of algorithm parameters has been
model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
both only require a single parameter: population size. \cite{posik:2011:parameterless}
leveraged these methods to create a fully parameter-less algorithm which was restricted
only to order-k fully decomposable noiseless problems.

@@

~\cite{bosman:2011:lsbbo} % LTGA with hill climbing

\subsection{P3}
Figure~\ref{fig-p3}.
\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber
    \If{solution $\notin hashset$}
      \State Add solution to $P_0$
      \State Add solution to $hashset$
    \EndIf

@@

Figure~\ref{fig-cluster-creation} Line~\ref{fig-cluster-creation-ordering}.

Figure~\ref{fig-cluster-usage} Line~\ref{fig-cluster-usage-donate}.

@@

>>>>

One area that has been very effective at reducing the number of algorithm parameters has been
model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
both only require a single parameter: population size. \cite{posik:2011:parameterless}
leveraged model building to create a fully parameter-less algorithm, but it is restricted to
only order-k fully decomposable noiseless problems.

@@

By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
the population size, and uses a fixed user specified size for each generation.

\subsection{Parameter-less Population Pyramid}
\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
performing optimization which did not require the user to provide any parameters. This is
achieved by combining efficient local search with the model building methods of LTGA using
an iteratively constructed hierarchy of populations.

The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
genetic algorithms, P3 does not follow a generational model. Instead, it maintains an iteratively
expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
This solution is brought to a local optimum using the algorithm in Figure~\ref{fig-hc}. If that
local optimum has not been added to the pyramid yet, the solution is added to the lowest
population $P_0$.

Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
and the number of populations stored increases over time. Initially the pyramid contains no solutions
or populations, meaning the user does not need to specify a population size.

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber
    \If{solution $\notin hashset$}
      \State Add solution to $P_0$
      \State Add solution to $hashset$
    \EndIf

@@

To accommodate P3's unique population structure, some of LTGA's clustering procedures needed
to be adapted. In LTGA
clusters are identified at the start of each generation, with those clusters used
to create all solutions produced in that generation. As P3 does not perform serial
generations, this process had to be modified. Instead, P3 updates the model each
time a solution is added to a population, the model is rebuilt. Furthermore, unlike
LTGA all solutions in the population are used to generate the model, not just the
winners of a binary tournament. This selection process is not necessary as the
worst solutions in the pyramid are already high quality due to local search.
Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
to provide no significant improvement.

Beyond the changes in population structuring, P3 also has slight modifications to
LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
from \emph{last merged first} ordering to \emph{smallest first} ordering.
This method applies gene clusters during crossover based on how many genes
are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
\cite{goldman:2012:ltga} found that this alternative was better at preserving
diversity, and therefore required smaller populations, but was not integrated into
our chosen canonical LTGA variant.

P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
Instead of choosing a single genetic donor for each cluster, P3 iterates over the
population in a random order until a solution in the population is found which
has a least one gene different for that cluster of genes than the improving solution.
This process increases the likelihood of an evaluation being performed for every cluster,
and helps test rare gene patterns in the population.

TODO Explain higher computational costs

@@

TODO Explain more than the sum of its parts
